Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-03-19 paper.qmd
Fixed Figure 7a

```{r}
#| label: hexbin-regular-two-curvy2
## hexagon binning to have regular hexagons
hb_obj_two_curvy2 <- hex_binning(
  data = tsne_two_curvy_scaled, 
  bin1 = 15, 
  r2 = r2)

a1_2 <- calc_bins_y(
  bin1 = 15, 
  r2 = r2
  )$a1

## Data set with all centroids
all_centroids_df2 <- hb_obj_two_curvy2$centroids

## Generate all coordinates of hexagons
hex_grid2 <- hb_obj_two_curvy2$hex_poly

## To obtain the standardise counts within hexbins
counts_df2 <- hb_obj_two_curvy2$std_cts
df_bin_centroids2 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df2, 
  counts_df = counts_df2) |>
  filter(drop_empty == FALSE) |>
  mutate(b1 = "b1 = 11")

hex_grid_with_counts_two_curvy2 <- full_join(
  hex_grid2, 
  df_bin_centroids2 |> select(hexID, std_counts), 
  by = c("hex_poly_id" = "hexID"))

umap_data_two_curvy2_with_hb_id <- hb_obj_two_curvy2$data_hb_id
df_all_two_curvy2 <- dplyr::bind_cols(training_data_two_curvy, umap_data_two_curvy2_with_hb_id)
df_bin_two_curvy2 <- avg_highd_data(data = df_all_two_curvy2, col_start = "x")

df_b_two_curvy2 <- df_bin_two_curvy2 |>
  dplyr::filter(hb_id %in% df_bin_centroids2$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_two_curvy2 <- df_b_two_curvy2[match(df_bin_centroids2$hexID, df_b_two_curvy2$hb_id),] |>
  dplyr::select(-hb_id)

@@

## Triangulate bin centroids
tr1_object_two_curvy2 <- tri_bin_centroids(
  df_bin_centroids2, x = "c_x", y = "c_y")
tr_from_to_df_two_curvy2 <- gen_edges(
  tri_object = tr1_object_two_curvy2)

@@

## Compute 2D distances
distance_two_curvy <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_two_curvy2, 
  start_x = "x_from", 
  start_y = "y_from", 
  end_x = "x_to", 
  end_y = "y_to", 
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_two_curvy <- find_lg_benchmark(
  distance_edges = distance_two_curvy, 
  distance_col = "distance")

```{r}
#| label: best-fit-two-curvy-proj

# Apply the scaling
df_model_data_two_curvy <- bind_rows(
  data_two_curvy, 
  true_model_two_curvy, 
  df_b_two_curvy2)

## First projection
projection <- cbind(
    c(0.11062,-0.02743,0.02786,-0.01930),
    c(-0.01110,-0.06886,0.04730,0.09890))

proj_obj1 <- get_projection(projection = projection, 
                            proj_scale = 1, 
                            scaled_data = scaled_gau_data, 
                            scaled_data_model = scaled_gau_data_model, 
                            distance_df_small_edges = distance_df_small_edges, 
                            axis_param = list(limits = 0.15, 
                                              axis_scaled = 5,
                                              axis_pos_x = -0.067, 
                                              axis_pos_y = -0.067, 
                                              threshold = 0.007))

# projection_scaled <- projection * 1
# 
# projected <- as.matrix(scaled_two_curvy_data) %*% projection_scaled
# 
# projected_df <- projected |>
#   tibble::as_tibble(.name_repair = "unique") |>
#   dplyr::rename(c("proj1" = "...1",
#                   "proj2" = "...2")) |>
#   dplyr::mutate(ID = dplyr::row_number()) 
# 
# ## For fitted model
# projected_model <- as.matrix(scaled_two_curvy_data_model) %*% projection_scaled
# 
# projected_model_df <- projected_model |>
#   tibble::as_tibble(.name_repair = "unique") |>
#   dplyr::rename(c("proj1" = "...1",
#                   "proj2" = "...2")) |>
#   dplyr::mutate(ID = dplyr::row_number())
# 
# model_df <- dplyr::left_join(
#   distance_df_small_edges_two_curvy2 |> select(-distance), 
#   projected_model_df, 
#   by = c("from" = "ID"))
# 
# names(model_df)[3:NCOL(model_df)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_from")
# 
# model_df <- dplyr::left_join(model_df, projected_model_df, by = c("to" = "ID"))
# names(model_df)[(2 + NCOL(projected_model_df)):NCOL(model_df)] <- paste0(names(projected_model_df)[-NCOL(projected_model_df)], "_to")

# axes_obj <- gen_axes(  
#   proj = projection * 5,
#   limits = 0.15,
#   axis_pos_x = -0.067,
#   axis_pos_y = -0.067,
#   axis_labels = names(scaled_two_curvy_data),
#   threshold = 0.007)
# 
# axes <- axes_obj$axes
# circle <- axes_obj$circle

two_curvy_proj_umap_model <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") + #31a354
  geom_point(
    size = 0.5,
    alpha = 0.2,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.1, 0.1)) + 
  ylim(c(-0.1, 0.1)) +
  interior_annotation("c2", sc_ltr_pos) +
  theme(
    legend.position = "none"
  )

two_curvy_proj_umap_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") + #31a354
  geom_point(
    size = 0.5,
    alpha = 0.2,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.1, 0.1)) + 
  ylim(c(-0.1, 0.1)) +
  interior_annotation("a3", sc_ltr_pos) +
  theme(
    legend.position = "none"
  )

two_curvy_proj_umap_all_model <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = true_model_df_proj, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#2b8cbe") +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") +
  geom_point(
    size = 0.5,
    alpha = 0.5,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.1, 0.1)) + 
  ylim(c(-0.1, 0.1)) +
  interior_annotation("a4", sc_ltr_pos) +
  theme(
    legend.position = "none"
  )

>>>>

```{r}
#| label: hexbin-regular-two-curvy2

num_bins_x_two_curvy <- 15

algo_obj_two_curvy2 <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_two_curvy, 
  nldr_data = tsne_two_curvy, 
  num_x_bins = num_bins_x_two_curvy,
  col_start = "x")

tsne_two_curvy_scaled <- algo_obj_two_curvy2$nldr_scaled
distance_two_curvy2 <- algo_obj_two_curvy2$distance_df
tr_from_to_df_two_curvy2 <- algo_obj_two_curvy2$tr_from_to_df
benchmark_two_curvy2 <- algo_obj_two_curvy2$benchmark
df_bin_centroids_two_curvy2 <- algo_obj_two_curvy2$df_bin_centroids
df_bin_two_curvy2 <- algo_obj_two_curvy2$df_bin
hex_grid_with_counts_two_curvy2 <- algo_obj_two_curvy2$hex_grid_with_counts

distance_df_small_edges_two_curvy2 <- distance_two_curvy2 |>
  filter(distance < benchmark_two_curvy2)

tr_from_to_df_two_curvy2 <- inner_join(
  tr_from_to_df_two_curvy2, distance_df_small_edges_two_curvy2, 
  by = c("from", "to"))

@@

@@
